Research Notes

### 2018-01-30 writups/notes.tex
01/28

\subsection{Nonturbulent Equations of Motion}

The equations of motion are the usual fluid equations
\begin{subequations}\label{se:fl_eq}
    \begin{align}
        \pd{\rho}{t} + \vec{\nabla} \cdot \rho \vec{u} &= 0,
            \label{eq:fl_eq.cont}\\
        \rd{\vec{u}}{t} &= -\frac{\vec{\nabla}P}{\rho} - g\hat{z},
            \label{eq:fl_eq.mom}
    \end{align}
\end{subequations}
where $\rd{}{t} = \pd{}{t} + \vec{u} \cdot \vec{\nabla}$ denotes the Lagrangian
derivative. We take $\vec{u}_0 = 0$ no background flow, $P_0(x, z) = P_0(z)
\propto e^{-z/H}$ where $H$ is the scale height of the atmosphere. We take the
fluid to be incompressible $\vec{\nabla} \cdot \vec{u} = 0$, which turns the
continuity equation \autoref{eq:fl_eq.cont} into $\rd{\rho}{t} = 0$. Finally, we
take the atmosphere to be isothermal, implying $\rho_0 \propto P_0 \propto
e^{-z/H}$, and more importantly $\frac{P_0}{\rho_0} = \frac{P_1}{\rho_1} =
\frac{k_BT}{\mu m_p}$. We denote $P_0 = P_0(z=0), \rho_0 = \rho_0(z=0)$.

We write out the new fluid equations under these assumptions; note that we do
not assume perturbed quantities are small and must keep all terms
\begin{subequations}\label{se:lin_eq}
    \begin{align}
        \pd{\rho_1}{t} &= \frac{u_{1z}}{H}\rho_0e^{-z/H} - \vec{u}_1 \cdot
            \vec{\nabla} \rho_1, \label{eq:lin_eq.cont}\\
        \pd{\vec{u}_1}{t} &=
            -\p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
            -\frac{-\frac{\hat{z}}{H}P_0e^{-z/H} + \frac{k_BT}{\mu m_p}
                \vec{\nabla} \rho_1}{\rho_0e^{-z/H} + \rho_1}
                - g\hat{z}, \label{eq:lin_eq.mom}.
    \end{align}
\end{subequations}
This consists of four equations in four unknowns, $\rho_1$ and the three
components of $\vec{u}_1$.

\subsection{Turbulent Dissipation}

In order to mimic the effect of turbulent dissipation, we introduce a viscosity
term into the equations of motion. Per the incompressible Navier-Stokes
equation, we quantify viscosity via $\nu \nabla^2 \vec{u}_1$ where $\nu$ is the
turbulent viscosity. For this problem, we simply use $\nu \neq
0.1\;\mathrm{cm^2/s}$ for air.

We identify the kinetic energy density carried by the perturbation to be
$\mathcal{U}_T = \frac{1}{2}\rho \p*{\vec{u}_1 \cdot \vec{u}_1}$. We require
$\rd{\mathcal{U}}{t} = 0$ under the dissipation-free fluid equations
\autoref{se:fl_eq} where we denote $\mathcal{U} \equiv \mathcal{U}_T +
\mathcal{U}_V$ the sum of the kinetic and potential energy densities; we assume
incompressibility as per our present problem:
\begin{align}
    \rd{\p*{\mathcal{U}_T + \mathcal{U}_V}}{t} = 0
        &= \frac{1}{2}u_1^2 \cancelto{0}{\rd{\rho}{t}}
        + \rho \vec{u}_1 \cdot \rd{\vec{u}_1}{t} + \rd{\mathcal{U}_V}{t}
        ,\nonumber\\
        &= \rho \p*{
                -\frac{\vec{u}_1 \cdot \vec{\nabla}P}{\rho} - u_{1z}g
            } + \rd{\mathcal{U}_V}{t},\nonumber\\
        &= -\cancelto{0}{\rd{P}{t}} + \cancelto{0}{\pd{P_0}{t}}
            + \pd{P_1}{t} - \rho u_{1z}g + \rd{\mathcal{U}_V}{t}.
\end{align}

%TODO figure this out from L&L, figure out the energy dissipation upon adding
%Navier-Stokes term.

@@

To set up the boundary condition at $z = 0$, we recall that gravity waves in an
atmosphere with $e^{-z/H}$ profile have form
\begin{equation}
    u_{1z} \propto e^{\frac{z}{2H}}e^{i\p*{k_xx + k_zz - \omega t}},
        \label{eq:bc_below}
\end{equation}
where ($N^2 \equiv \frac{g}{H}$ is the Brunt-V\"ais\"al\"a frequency in an
incompressible, stratified atmosphere)
\begin{equation}
    \omega^2 = \frac{N^2 k_x^2}{k_x^2 + k_z^2 + \frac{1}{4H^2}}.
        \label{eq:omega}
\end{equation}
Thus, at constant $z = 0$ we must have
\begin{equation}
    \vec{u}_{1z}(z = 0) = e^{i(k_xx - \omega t)}\label{eq:bc_u1z}.
\end{equation}

@@

For the boundary conditions on $\rho$, we note that in the linear regime it
should just have a phase offset from $u_{1z}$, thus we choose $\rho(z = 0)
\propto e^{\frac{z}{2H}}ie^{i\p*{k_xx + k_zz - \omega t}}$ and a similar
treatment at $z = L_z$, taking both a Dirichlet and Neumann BC\@. This should be
sufficient information to constrain our problem, as satisfying incompressibility
along the boundaries constrains $u_{1x}$.

\subsubsection{Old Stuff---$u_{1z}(z = L_z)$}

\textbf{Note:} The reason the below is difficult to make work is that it tries
to restore a linear treatment to the wave. However, since we must capture
nonlinear dynamics, we lose the concept of the ``phase'' of the wave in the wave
breaking region, and it proves difficult to write down a BC that is
phase-independent.

However, at any finite boundary there is nonzero reflection thanks to the
continually varying impedance. \emph{The correct BC to use when performing a
finite truncation must match propagation characteristics with an untruncated
medium.}

We assume the perturbations have once again become small at $L_z$, i.e.\
dissipation between $[0, L_z]$ is sufficient that only a small amplitude
perturbation persists to $z = L_z$. This is a necessary criterion for choosing
$L_z$ else we do not have a representative portrait of the energy dissipation
profile of the propagating wave with our simulation. Then we recall that in the
linear regime, the solution given by \autoref{eq:bc_below} is the propagating
solution. We assume that $k_x, \omega$ are the same as the values at the $z = 0$
boundary, then
\begin{equation}
    \frac{\rho_{,z}}{\rho} = \frac{u_{1z, z}}{u_{1z}} = \frac{1}{2H} + ik_z.
        \label{eq:bc_above}
\end{equation}
Recall that since our equations of motion are nonlinear, we cannot use complex
notation except in very limited, time-averaged cases. Thus, how to implement
this BC is tricky\dots

>>>>

The fluid equations are
\begin{subequations}\label{se:fl_eq}
    \begin{align}
        \pd{\rho}{t} + \vec{\nabla} \cdot \rho \vec{u} &= 0,
            \label{eq:fl_eq.cont}\\
        \rd{\vec{u}}{t} &= -\vec{\nabla}\frac{P}{\rho} - g\hat{z},
            \label{eq:fl_eq.mom}
    \end{align}
\end{subequations}
where we will take gravity to be uniform throughout the domain of interest. We
will study for no background flow $\vec{u}_0 = 0$ and in vertically stratified
atmosphere $\rho_0 \propto e^{-z/H}$. In the absence of any perturbations it is
then easy to show that $P_0 = \rho_0 g H$.

\subsection{Linear Regime, Incompressible}

First, we solve the incompressible case $c_s^2 \to \infty, \vec{\nabla} \cdot
\vec{u} = 0$ in the linear regime. The fluid equations to first order reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\rho_0 &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - P_1 \vec{\nabla}\frac{1}{\rho_0}
            + \rho_1 gH \vec{\nabla}\frac{1}{\rho_0}
            + \frac{gH}{\rho_0} \vec{\nabla}\rho_1.
    \end{split}\label{eq:lin_incomp}
\end{equation}
Substituting in plane waves of form $e^{i\p*{\vec{k} \cdot \vec{r} - \omega t}}$
and the given vertically stratified $e^{-z/H}$ density and pressure profiles
gives system of equations
\begin{align}
    \begin{bmatrix}
        -i\omega & 0 & -\frac{\rho_0}{H} & 0\\
        0 & ik_x & ik_z & 0\\
        0 & -i\omega & 0 & \frac{ik_x}{\rho_0}\\
        \frac{P_0}{\rho_0^2H} & 0 & -i\omega & \frac{ik_z}{\rho_0}
    \end{bmatrix}\begin{bmatrix}
        \rho_0 \\ u_{1x} \\ u_{1z} \\ P_1
    \end{bmatrix} &= 0,\nonumber\\
    \omega^2 k^2 + \frac{P_0}{\rho_0^2H}k_x^2 = \omega^2k^2 - N^2k_x^2 &= 0.
\end{align}
We identify $\frac{P_0}{\rho_0^2H} = \frac{g}{H} = N^2$ the Brunt-V\"ais\"al\"a
frequency. This agrees with the usual incompressible dispersion relation
$\omega^2 = N^2\cos^2\theta$ where $\theta$ is the angle formed between
$\vec{k}$ and $\hat{x}$.

\subsection{Linear Regime, Compressible}

@@

To set up the boundary condition at $z = 0$, we recall that gravity waves in an
atmosphere with $e^{-z/H}$ profile have form
\begin{equation}
    u_{1z} \propto e^{\frac{z}{2H}}e^{i\p*{k_xx + k_zz - \omega t}},
        \label{eq:bc_below}
\end{equation}
where ($N^2 \equiv \frac{g}{H}$ is the Brunt-V\"ais\"al\"a frequency in an
incompressible, stratified atmosphere)
\begin{equation}
    \omega^2 = \frac{N^2 k_x^2}{k_x^2 + k_z^2 + \frac{1}{4H^2}}.
        \label{eq:omega}
\end{equation}
Thus, at constant $z = 0$ we must have
\begin{equation}
    \vec{u}_{1z}(z = 0) \propto e^{i(k_xx - \omega t)}\label{eq:bc_u1z}.
\end{equation}

@@

For the boundary conditions on $\rho$, we note that in the linear regime it
should just have a phase offset from $u_{1z}$, thus we choose $\rho(z = 0)
\propto e^{\frac{z}{2H}}ie^{i\p*{k_xx + k_zz - \omega t}}$ and a similar
treatment at $z = L_z$, taking both a Dirichlet and Neumann BC\@.
